---
title: "Calculus with R"
author: "Daniel Kaplan"
date: "October 17, 2017"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Calculus_with_R}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r include = FALSE}
library(mosaic)
library(mosaicData)
library(mosaicCalc)
library(ggformula)
```


The `mosaicCalc` package provides basic functions for computing derivatives and anti-derivatives, as well as for integrating ordinary differential equations. It was designed for teaching calculus. 

`mosaicCalc` was designed along with the `mosaic` package. Graphics capabilities for displaying functions built with `mosaicCalc` are provided by `mosaic`, `ggformula`, and related packages.

Install `mosaicCalc` CRAN using the usual method. The GitHub repository for the package is `ProjectMOSAIC/mosaicCalc` and the latest patch can be installed from there using `devtools::install_github()`.

When using `mosaicCalc`, typically, you will also want to work with `mosaic`. So load both packages:

```{r}
library(mosaic)
library(mosaicCalc)
```

## Creating functions

Calculus operations such differentiation and integration take a function as input and produce a function as output. There are many ways to create functions. For simplicity, our initial examples will be functions $f(x)$ with a single input $x$.

Many R users are familiar with the `function` keyword in R which enables functions to be defined, e.g. this simple quadratic function:

```{r}
f1 <- function(x) 1 - x^2
```

The `mosaicCalc` package provides a more concise way to define functions that uses the formula syntax in R.
```{r}
f2 <- makeFun(1 - x ^ 2 ~ x)
```
The formula tilde (`~`) is typographically small, but it plays an essential role in `mosaicCalc` functions. The function body goes to the left of tilde while the formal variables are to the right.

You can include symbolic parameters in functions by using a symbolic entity in the body of the function, like this:

```{r warning = FALSE}
f3 <- makeFun(a - b * x ^ 2 ~ x)
```

The `makeFun()` function does not formally distinguish between arguments (e.g. $x$) and parameters (e.g. `a` and `b`). As a result, `f3` will have three arguments:

```{r}
f3
```

You can also include the parameters explicitly to the right of the tilde. And you can assign default numerical values to the parameters by adding additional named arguments. For example:
```{r warning = FALSE}
f4 <- makeFun(a - b * x ^2 ~ x & a & b, a = 1, b = 1)
```

Using `makeFun()` rather than `function()` is to be preferred
