---
title: "Calculus with R"
author: "Daniel Kaplan"
date: "May 6, 2020"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Calculus_with_R}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r include = FALSE}
library(mosaic)
library(mosaicData)
library(mosaicCalc)
library(ggformula)
```

This document is intended to guide students or instructors who are beginners with R and who want to learn how to perform calculus operations. The document describes the use of the "R package" called `{mosaicCalc}`. The package provides software with a unified and consistent user interface for the calculus operations. It was developed specifically to support the [*MOSAIC Calculus*](dtkaplan.github.io/MC2) textbook, available freely online. However, instructors and other textbooks can and do use `{mosaicCalc}` independently of *MOSAIC Calculus*.

R, as you may know, is an environment and language for technical computing. It's most widely associated with data science and statistics, but it is also well suited to calculus. R is widely used in many diverse fields. This results, in part, from the "open source" nature of the R environment. An aspect of "open source" is that anyone can contribute new capabilities to the environment. Those new capabilities are typically provided in a officially standardized format called an "R package." There are thousands of such packages available through the official distribution network, called CRAN. `{mosaicCalc}` is one of those packages and works in conjunction with other CRAN packages.

As you can see from the previous two paragraphs, we are using a `monospaced font` when referring to names that are part of the computer system. Many of the names you will encounter, like `D()`, `antiD()`, and `Integrate()` will be names of "functions" that allow you to perform computations such as differentiation and integration. `{mosaicCalc}` is not a function. Rather it is a *collection* of functions that inter-operate and follow similar conventions. The written for name is written inside curly braces as a reminder to the human reader that `{mosaicCalc}` is a collection of functions. 

To use `{mosaicCalc}` successfully, you need to learn 

- a small set of simple **object types**, such as "tilde expressions," "functions," and "bounds."
- a few **functions** that carry out calculus and other mathematical operations, graphing, etc.
- how to construct **commands** that tell the R environment to set up and carry out the operations used to apply calculus to the settings of interest in your chosen field.

## Tilde expressions


## Functions


## Bounds


## Graphing





The `{mosaicCalc}` package provides R functions for computing derivatives and anti-derivatives, graphics, zero-finding (often called "solving"), numerical optimization, and integrating ordinary differential equations. It is designed for teaching calculus. 



`mosaicCalc` extends the `mosaic` package. Graphics capabilities for displaying functions built with `mosaicCalc` are provided by `mosaic`, `ggformula`, and related packages.

Install `mosaicCalc` CRAN using the usual method. The GitHub repository for the package is `github.com/ProjectMOSAIC/mosaicCalc` and the latest patch can be installed from there using `remotes::install_github("ProjectMOSAIC/mosaicCalc)`.

When using `mosaicCalc`, typically, you will also want to work with `mosaic`. So load both packages:

```{r}
library(mosaic)
library(mosaicCalc)
```

## Calculus, computing, and notation

A dominant style of notation in teaching calculus is the analytic style, where functions are written in the form $f = 1 / x$, derivatives are referred to in any of several notations such as $f'$ or $\frac{d}{dx} f$. Integrals are denoted $\int f(x) dx$. In this notation, $dx$ is a formal device that serves to identify the variable with respect to which the operation of integration or differentiation is being performed.

This contrasts with mainstream computer languages where functions are identified by *names* and there is no  universal notation for "with respect to ....". The calculus operations in `mosaicCalc` exploit the tilde-formula notation in R, e.g.  `1 / x ~ x`. The symbol on the right-hand side of tilde identifies the variable with respect to which the operation is to be performed. To the left of tilde is a statement in standard computer notation of the transformation carried out by the function.

Traditionally, a typographic distinction is made between "variables" and "parameters." For instance, it's implicit that the straight-line function  $mx + b$ involves the variable $x$ and the parameters $m$ and $b$. Usually the *variables* are letters toward the end of the alphabet such  as $t$, $x$,  and $y$ whereas *parameters* are written using earlier letters in the alphabet (e.g., $a$, $b$, $c$, $k$) sometimes using upper-case letters,  Greek letters, and/or subscripts. 

In `mosaicCalc`, no formal distinction is made between variables and parameters except as provided by the "with respect to ..." component of a formula. In `mosaicCalc` you would write `m * x + b ~ x` as a formula for a straight-line function. 

A formula like `m * x + b ~ x` stands for a function, but it is not an R function. Formulas in `mosaicCalc` are routinely turned into actual R functions. This happens automatically in operations such as differentiation and anti-differentiation, which return a function rather than a formula.

The operation `makeFun()` in `mosaicCalc` performs this action. For instance, here is the definition of a function called `f`:

```{r}
f <- makeFun(m * x + b ~ x)
f
```

Note that `f` is a function of *three* arguments. There's no formal distinction  being made between the "variable" and the "parameters." (`makeFun()` can also be used to turn linear and logistic regression reports into functions.)

Contrast this with the way an R programmer might naturally define the straight-line function:

```{r}
# NOT the style in mosaicCalc
straight_line <- function(x) m * x + b
```

`straight_line` is a legitimate function in R, but the parameters $m$ and $b$ are not represented by arguments to the function. Instead, they are set by the scoping rules of the R language. In particular, given this definition  of `straight_line`, at the time that `straight_line()` is applied to an input, the values of `m` and `b` will be looked up in the R environment in which the function was defined.

The `mosaicCalc` approach is to avoid as much as possible the need to understand the lexical scoping used by R. And so *every* quantity in a function made by the likes of `makeFun()` will become an argument. 

In `makeFun()` the "with respect to ..." component of the formula is used only to determine the order of the arguments: symbols used on the right-hand side of the tilde are put first in the argument list.

## Differentiation and anti-differentiation

The `mosaicCalc` operations for differentiation and anti-differentiation are called `D()` and `antiD()` respectively. Like `makeFun()`, `D()` and `antiD()` take as a primary argument a *formula* where the right-hand side identifies the "with respect to ..." variable. And, all three of those operations are similar in that they return a *function* (not a formula) whose arguments include  every symbol used in the  body  of the function.

```{r}
D(a / x ~ x)
```

In all three operations, you  can declare a default value for the parameter, which enables you have the parameter symbolically identified while not having to assign a value  to the parameter every time it is evaluated. For instance,

```{r}
makeFun(a / x ~ x, a = 3)
D( a / x ~ x, a = 3)
```

Famously, anti-differentiation introduces an additive quantity usually denoted C, so $\int 1/x\ dx \rightarrow \ln x + C$. This $C$ is treated as a parameter in the output of `mosaicCalc::antiD()`:

```{r}
antiD(a / x ~ x)
```

## Analytic or numeric

The symbolic capabilities of `mosaicCalc` are built on top of the  `{Deriv}` and `{Ryacas}` packages. The interface uses the tilde-expression format; the translation of this into a suitable format for `{Deriv}` and `{Ryacas}` is handled automatically and is transparent to the user. 

```{r}
D(a * x + b * x^2 + c * sin(x^2) ~ x)
```

while `antiD()` will generate a numeric-integration result:


```{r}
F <- antiD(a * x + b * x^2 + c * sin(x^2) ~ x)
F
```

Note that even though the function `F()` involves applying numerical methods, it is still a function and can itself be differentiated or integrated using `D()` and `antiD()` in `mosaicCalc`.

## Graphics

The meaning of "with respect to ..." when making a graphic is to set the coordinate axes. For instance, the `mosaic` package includes a function `plotFun()`  

```{r}
mosaic::plotFun(1 / x ~ x, xlim=c(0,5))
```

`plotFun()` can be thought of as a legacy function. It uses the `lattice` graphics system.  (We plan soon to be adding  `ggplot2` compatible calculus graphics functions.)

## 
